---
title: "NXS TCS Article — Chapter Analysis"
subtitle: "How close are the assessment chapters to the keypaper concepts?"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    grid:
        sidebar-width: 0px
        body-width: 4000px
        margin-width: 200px
        gutter-width: 1.5rem
execute:
  message: false
  warning: false
  cache: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(htmltools)
library(knitr)
```

This report measures **how similar the papers in each assessment chapter are to
the keypaper concepts** (the TCA/Nexus definitions). It is driven by the
combined per-chapter scores (`scores_combined`): each corpus work scored from
its `title_abstract` embedding, or its `title` embedding where it has no
abstract, against the keypaper set — one row per `(work, assessment, chapter)`.

A work tagged with several Zotero chapters is counted **once per chapter**
(honest for "which chapters supply the matches"). Chapter labels are shown as
tagged, except the malformed NXS tag `Chapter(s) 5`, relabelled to `Chapter 5`;
the `Chapter 5.0…5.6` sub-chapters are kept separate.

# Overall: which chapters are closest to the keypapers?

Median (with inter-quartile bar) of each work's maximum cosine similarity to
its nearest keypaper, per assessment/chapter. Higher = the chapter's papers sit
closer to the keypaper concepts overall.

```{r}
#| label: fig-chapter-alignment
#| fig-cap: "Chapter alignment to the keypaper set: median (bar) and bootstrap 95% CI of the median (error bar) of per-work max cosine similarity to the nearest keypaper, per assessment/chapter, sorted ascending. Chapters whose CIs do not overlap are really different; small-n chapters get wide CIs. n = works × chapter memberships in that chapter."
#| fig-height: 8
tar_read(viz_chapter_alignment_fig)
```

## Which chapters are *really* different? (Cliff's δ)

Raw cosine gaps are hard to read on the compressed similarity scale, and with
thousands of works trivial gaps look "significant". Cliff's δ is an effect
size: for a pair of chapters it is P(a row-chapter work scores higher) −
P(it scores lower). **0 ≈ indistinguishable**; |δ| ≈ 0.15 / 0.33 / 0.47 is a
small / medium / large separation. Read a row against the columns: red = the
row chapter's papers tend to score *higher* than the column's, blue = lower.

```{r}
#| label: fig-chapter-cliffs-delta
#| fig-cap: "Pairwise Cliff's δ between assessment/chapters on per-work max keypaper similarity. Rows/columns ordered by median alignment (highest first). δ near 0 (white) = the two chapters' similarity distributions are indistinguishable."
#| fig-height: 9
tar_read(viz_chapter_cliffs_delta_fig)
```

# In detail: chapter × keypaper

Median similarity of each assessment/chapter to each **individual** keypaper.
Bright cells = the chapter's papers are close to that concept; dark rows =
concepts the corpus has little near (orphans); a chapter's brightest cells are
the concepts it most aligns with. Rows (keypapers) and columns (chapters) are
ordered by overall median.

```{r}
#| label: fig-chapter-keypaper-heatmap
#| fig-cap: "Median cosine similarity per (assessment/chapter, keypaper). Rows = keypaper concepts, columns = assessment/chapter."
#| fig-height: 12
tar_read(viz_chapter_keypaper_heatmap_fig)
```

# Top matches per keypaper, by chapter

Pick a keypaper concept to browse its highest-scoring corpus matches. The
histogram stacks the matches' similarity by **assessment/chapter**, so you can
see which chapters supply a concept's closest papers. **Toggle chapters in the
legend** (click to show/hide, double-click to isolate); `No Chapter` works are
**hidden by default** — click their legend entry to include them. The table
below carries `assessment` and `chapter` columns (filterable), and the CSV
button exports the currently filtered rows. Links go to the DOI when available,
otherwise the OpenAlex page.

```{r}
#| label: tbl-top-matches-combined
tar_read(tbl_top_matches_combined)
```

# Keypaper relevance across the corpus

The sections above break similarity down by chapter. These look at the
keypaper-similarity signal across the **whole corpus** (all works, per
embedding variant) — context for reading the per-chapter results.

For each corpus work, the **max** cosine similarity against any keypaper is the
single most useful summary — "how close is this work to its nearest reference?"
A well-calibrated embedding produces a bimodal distribution: relevant works
pile up on the right, off-topic on the left.

```{r}
#| label: tbl-score-summary
tar_read(viz_score_summary_tbl) |>
  knitr::kable(
    caption = "Summary statistics of the embeddings, split by source of embedding.",
    format.args = list(big.mark = ",")
  )
```

```{r}
#| label: fig-score-dist
#| fig-cap: "Distribution of max cosine similarity to keypapers, per variant. Higher = closer to nearest reference. A clear bimodal shape indicates the embedding separates topic-relevant from off-topic works."
tar_read(viz_score_dist_fig)
```

```{r}
#| label: fig-score-ecdf
#| fig-cap: "Empirical CDF of max cosine similarity to nearest keypaper, per variant (subsample of 50,000 works per variant, seed = 13). Read horizontally: for any score threshold, the curves give the fraction of works at or below that score."
tar_read(viz_score_ecdf_fig)
```

## Relevance threshold sweep

How many corpus works count as "relevant" as the similarity threshold rises —
useful for picking a cut-off when filtering to keypaper-relevant works.

```{r}
#| label: fig-threshold
#| fig-cap: "Number (and fraction) of corpus works with max keypaper similarity at or above each threshold, per variant. A gentle slope means the cut-off choice matters little; a cliff marks a natural relevance boundary."
tar_read(viz_threshold_fig)
```

## Score vs publication year

Does similarity to keypapers correlate with how recent a work is? A strong bias
either way suggests the keypaper set is anchored to a particular era.

```{r}
#| label: fig-score-year
#| fig-cap: "2D histogram of max cosine similarity (y) against publication year (x), per variant. Log-scale fill reveals the tail; uniform vertical density at every year is the null-hypothesis appearance — no temporal bias."
tar_read(viz_score_year_fig)
```

## Score vs citation count

Are high-similarity works also highly cited? Topical relevance and citation
impact are different signals — a strong correlation would suggest the embedding
is picking up "prestige", not topic.

```{r}
#| label: fig-citation-score
#| fig-cap: "2D histogram of max cosine similarity (y) against log10(cited_by_count + 1) (x), per variant. A horizontal band = no relationship (good); a diagonal would suggest the embedding co-varies with citation impact rather than topic."
tar_read(viz_citation_score_fig)
```

## Score by work type

Does the embedding score some OpenAlex work types (articles, reviews, book
chapters, …) as more keypaper-relevant than others?

```{r}
#| label: fig-type-score-heatmap
#| fig-cap: "Median max-similarity score per (type, variant), for types with ≥ 0.5% of the corpus. Cells annotated with `median (n=…)`. Types sorted by overall median."
tar_read(viz_type_score_heatmap_fig)
```

```{r}
#| label: fig-type-score-box
#| fig-cap: "Score distribution per type, per variant. Boxes show q05–q25–median–q75–q95 (precomputed). Filtered to types with ≥ 0.5% of the corpus."
tar_read(viz_type_score_box_fig)
```

## Top / bottom matches per variant

Sanity check: do the highest-scoring works read as on-topic, and the lowest as
off-topic? If not, either the embedding isn't capturing the topic or the
keypaper set is too narrow.

```{r}
#| label: tbl-top-bottom
#| results: asis
tb <- tar_read(viz_top_bottom)
for (v in names(tb)) {
  cat("###", v, "\n\n")
  print(knitr::kable(tb[[v]]))
  cat("\n\n")
}
```

# Keypaper set

## Keypaper coherence

How tight is the keypaper set itself? Pairwise cosine similarity between the
keypapers (title_abstract variant), reordered by average-linkage hierarchical
clustering. A coherent set shows clear high-similarity blocks on the diagonal.

```{r}
#| label: fig-keypaper-self-sim
#| fig-cap: "Pairwise cosine similarity between keypapers (title_abstract variant). Axis order is from hierarchical clustering (1 − cosine, average linkage). Block structure on the diagonal indicates topical sub-groups within the keypaper set."
tar_read(viz_keypaper_self_sim_fig)
```

## Per-keypaper score distribution

For each of the 62 keypapers, the distribution of cosine similarity across the
whole corpus. A high median = the keypaper attracts many near matches (possibly
too generic); a low median = orphaned (the corpus has nothing close). Sorted by
median, descending.

```{r}
#| label: fig-keypaper-score-dist
htmltools::div(
  style = "height: 3200px; overflow: visible;",
  tar_read(viz_keypaper_score_dist_fig)
)
```

*Per-keypaper score distribution (q05–q25–median–q75–q95). Top = high across the
corpus (broad/generic concepts); bottom = orphans (topic not well-represented).
Inputs to keypaper-set curation.*
