---
title: "NXS TCS Article — Chapter Analysis"
subtitle: "How close are the assessment chapters to the keypaper concepts?"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    grid:
        sidebar-width: 0px
        body-width: 4000px
        margin-width: 200px
        gutter-width: 1.5rem
execute:
  message: false
  warning: false
  cache: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(htmltools)
library(knitr)
```

This report measures **how similar the papers in each assessment chapter are to
the keypaper concepts** (the TCA/Nexus definitions). It is driven by the
combined per-chapter scores (`scores_combined`): each corpus work scored from
its `title_abstract` embedding, or its `title` embedding where it has no
abstract, against the keypaper set — one row per `(work, assessment, chapter)`.

A work tagged with several Zotero chapters is counted **once per chapter**
(honest for "which chapters supply the matches"). Chapter labels are shown as
tagged, except the malformed NXS tag `Chapter(s) 5`, relabelled to `Chapter 5`;
the `Chapter 5.0…5.6` sub-chapters are kept separate.

# Overall: which chapters are closest to the keypapers?

Median (with inter-quartile bar) of each work's maximum cosine similarity to
its nearest keypaper, per assessment/chapter. Higher = the chapter's papers sit
closer to the keypaper concepts overall.

```{r}
#| label: fig-chapter-alignment
#| fig-cap: "Chapter alignment to the keypaper set: median (bar) and bootstrap 95% CI of the median (error bar) of per-work max cosine similarity to the nearest keypaper, per assessment/chapter, sorted ascending. Chapters whose CIs do not overlap are really different; small-n chapters get wide CIs. n = works × chapter memberships in that chapter."
#| fig-height: 8
tar_read(viz_chapter_alignment_fig)
```

## Which chapters are *really* different? (Cliff's δ)

Raw cosine gaps are hard to read on the compressed similarity scale, and with
thousands of works trivial gaps look "significant". Cliff's δ is an effect
size: for a pair of chapters it is P(a row-chapter work scores higher) −
P(it scores lower). **0 ≈ indistinguishable**; |δ| ≈ 0.15 / 0.33 / 0.47 is a
small / medium / large separation. Read a row against the columns: red = the
row chapter's papers tend to score *higher* than the column's, blue = lower.

```{r}
#| label: fig-chapter-cliffs-delta
#| fig-cap: "Pairwise Cliff's δ between assessment/chapters on per-work max keypaper similarity. Rows/columns ordered by median alignment (highest first). δ near 0 (white) = the two chapters' similarity distributions are indistinguishable."
#| fig-height: 9
tar_read(viz_chapter_cliffs_delta_fig)
```

The chart above collapses each work to its single best-matching keypaper
(`max_sim`) before comparing chapters. The interactive version below lets you
pick, from the combobox, either that same collapsed view or any **individual**
keypaper's raw similarity — i.e. "which chapters differ on their match to
*this specific concept*" rather than "on their best match to any concept".
Black borders mark |δ| ≥ 0.147 (a "real", non-negligible difference) in
either view.

```{r}
#| label: fig-chapter-cliffs-delta-interactive
#| fig-cap: "Interactive pairwise Cliff's δ between assessment/chapters. Use the combobox to switch between the collapsed max-keypaper-similarity view and any individual keypaper's raw similarity. Black borders mark |δ| ≥ 0.147 (a real, non-negligible difference)."
#| fig-height: 10
tar_read(viz_chapter_cliffs_delta_interactive_fig)
```

# In detail: chapter × keypaper

Median similarity of each assessment/chapter to each **individual** keypaper.
Bright cells = the chapter's papers are close to that concept; dark rows =
concepts the corpus has little near (orphans); a chapter's brightest cells are
the concepts it most aligns with. Rows (keypapers) and columns (chapters) are
ordered by overall median. With 161 keypapers this is too long to read as one
static chart — use the **keyset checkboxes** below to isolate one or two
keysets at a time; row labels are truncated to 30 characters, full titles are
on hover.

```{r}
#| label: fig-chapter-keypaper-heatmap-interactive
#| fig-cap: "Median cosine similarity per (assessment/chapter, keypaper), filterable by keyset. Rows = keypaper concepts, columns = assessment/chapter."
#| fig-height: 12
tar_read(viz_chapter_keypaper_heatmap_interactive_fig)
```

# Top matches per keypaper, by chapter

Pick a keypaper concept to browse its highest-scoring corpus matches. The
histogram stacks the matches' similarity by **assessment/chapter**, so you can
see which chapters supply a concept's closest papers. **Toggle chapters in the
legend** (click to show/hide, double-click to isolate); `No Chapter` works are
**hidden by default** — click their legend entry to include them. The table
below carries `assessment` and `chapter` columns (filterable), and the CSV
button exports the currently filtered rows. Links go to the DOI when available,
otherwise the OpenAlex page.

```{r}
#| label: tbl-top-matches-combined
tar_read(tbl_top_matches_combined)
```

# Keypaper relevance across the corpus

The sections above break similarity down by chapter. These look at the
keypaper-similarity signal across the **whole corpus** (all works, per
embedding variant) — context for reading the per-chapter results.

For each corpus work, the **max** cosine similarity against any keypaper is the
single most useful summary — "how close is this work to its nearest reference?"
A well-calibrated embedding produces a bimodal distribution: relevant works
pile up on the right, off-topic on the left.

```{r}
#| label: tbl-score-summary
tar_read(viz_score_summary_tbl) |>
  knitr::kable(
    caption = "Summary statistics of the embeddings, split by source of embedding.",
    format.args = list(big.mark = ",")
  )
```

```{r}
#| label: fig-score-dist
#| fig-cap: "Distribution of max cosine similarity to keypapers, per variant. Higher = closer to nearest reference. A clear bimodal shape indicates the embedding separates topic-relevant from off-topic works."
tar_read(viz_score_dist_fig)
```

```{r}
#| label: fig-score-ecdf
#| fig-cap: "Empirical CDF of max cosine similarity to nearest keypaper, per variant (subsample of 50,000 works per variant, seed = 13). Read horizontally: for any score threshold, the curves give the fraction of works at or below that score."
tar_read(viz_score_ecdf_fig)
```

## Relevance threshold sweep

How many corpus works count as "relevant" as the similarity threshold rises —
useful for picking a cut-off when filtering to keypaper-relevant works.

```{r}
#| label: fig-threshold
#| fig-cap: "Number (and fraction) of corpus works with max keypaper similarity at or above each threshold, per variant. A gentle slope means the cut-off choice matters little; a cliff marks a natural relevance boundary."
tar_read(viz_threshold_fig)
```

## Score vs publication year

Does similarity to keypapers correlate with how recent a work is? A strong bias
either way suggests the keypaper set is anchored to a particular era.

```{r}
#| label: fig-score-year
#| fig-cap: "2D histogram of max cosine similarity (y) against publication year (x), per variant. Log-scale fill reveals the tail; uniform vertical density at every year is the null-hypothesis appearance — no temporal bias."
tar_read(viz_score_year_fig)
```

## Score vs citation count

Are high-similarity works also highly cited? Topical relevance and citation
impact are different signals — a strong correlation would suggest the embedding
is picking up "prestige", not topic.

```{r}
#| label: fig-citation-score
#| fig-cap: "2D histogram of max cosine similarity (y) against log10(cited_by_count + 1) (x), per variant. A horizontal band = no relationship (good); a diagonal would suggest the embedding co-varies with citation impact rather than topic."
tar_read(viz_citation_score_fig)
```

## Score by work type

Does the embedding score some OpenAlex work types (articles, reviews, book
chapters, …) as more keypaper-relevant than others?

```{r}
#| label: fig-type-score-heatmap
#| fig-cap: "Median max-similarity score per (type, variant), for types with ≥ 0.5% of the corpus. Cells annotated with `median (n=…)`. Types sorted by overall median."
tar_read(viz_type_score_heatmap_fig)
```

```{r}
#| label: fig-type-score-box
#| fig-cap: "Score distribution per type, per variant. Boxes show q05–q25–median–q75–q95 (precomputed). Filtered to types with ≥ 0.5% of the corpus."
tar_read(viz_type_score_box_fig)
```

## Top / bottom matches per variant

Sanity check: do the highest-scoring works read as on-topic, and the lowest as
off-topic? If not, either the embedding isn't capturing the topic or the
keypaper set is too narrow.

```{r}
#| label: tbl-top-bottom
#| results: asis
tb <- tar_read(viz_top_bottom)
for (v in names(tb)) {
  cat("###", v, "\n\n")
  print(knitr::kable(tb[[v]]))
  cat("\n\n")
}
```

# Keypaper set

## Keypaper coherence

How tight is the keypaper set itself? Pairwise cosine similarity between the
keypapers (title_abstract variant), reordered by average-linkage hierarchical
clustering. A coherent set shows clear high-similarity blocks on the diagonal.

```{r}
#| label: fig-keypaper-self-sim
#| fig-cap: "Pairwise cosine similarity between keypapers (title_abstract variant). Axis order is from hierarchical clustering (1 − cosine, average linkage). Block structure on the diagonal indicates topical sub-groups within the keypaper set."
tar_read(viz_keypaper_self_sim_fig)
```

## Per-keypaper score distribution

For each of the 62 keypapers, the distribution of cosine similarity across the
whole corpus. A high median = the keypaper attracts many near matches (possibly
too generic); a low median = orphaned (the corpus has nothing close). Sorted by
median, descending.

```{r}
#| label: fig-keypaper-score-dist
htmltools::div(
  style = "height: 3200px; overflow: visible;",
  tar_read(viz_keypaper_score_dist_fig)
)
```

*Per-keypaper score distribution (q05–q25–median–q75–q95). Top = high across the
corpus (broad/generic concepts); bottom = orphans (topic not well-represented).
Inputs to keypaper-set curation.*

# Keyset linkage

How do the concept definitions link *to one another* across the four key-sets
(the worksheet groups: `TCA_Annex_3_3`, `TCA_Approaches_3_2`,
`TCA_Actions_Ch5`, `Nexus_Response_Options`)? Two directed links are of
particular interest — **TCA Action → Nexus Response Option** and
**TCA Approaches → TCA Action** — each shown below as a set of Sankeys (top 3
targets per source node). Three complementary similarity signals drive them:

- **Stage 1 — definition-text similarity (embeddings):** cosine similarity
  between the keypaper definition embeddings (title_abstract, SPECTER2). Two
  definitions link if their *wording/concepts* are close.
- **Stage 2 — cited-literature overlap (no embeddings):** two definitions link
  if they *cite the same works* — Jaccard overlap of the resolved cited-work
  sets (only pairs sharing ≥1 work).
- **Stage 3 — cited-literature embedding similarity:** two definitions link if
  the *literature they cite* is semantically close — mean pairwise cosine
  between the two definitions' cited-work embedding sets (computed as the dot
  product of their unit-vector centroids).

## Keyset × keyset overview

Each Sankey below isolates one source → target keyset pair. These matrices give
the whole picture: the mean similarity between *every* pair of key-sets, per
signal stage (read a row against the columns).

```{r}
#| label: fig-keyset-matrix-stage1
#| fig-cap: "Stage 1 — keyset × keyset mean definition-embedding cosine similarity."
tar_read(viz_keyset_matrix_stage1_fig)
```

```{r}
#| label: fig-keyset-matrix-stage2
#| fig-cap: "Stage 2 — keyset × keyset mean Jaccard overlap of cited-work sets."
tar_read(viz_keyset_matrix_stage2_fig)
```

```{r}
#| label: fig-keyset-matrix-stage3
#| fig-cap: "Stage 3 — keyset × keyset mean cited-literature embedding similarity."
tar_read(viz_keyset_matrix_stage3_fig)
```

## TCA Action ↔ Nexus Response Options

Directed **TCA Action → Nexus Response Option**: for each Action, its top 3
most-similar Response Options under each signal stage.

### Stage 1 — definition-text similarity (embeddings)

```{r}
#| label: fig-sankey-act-resp-stage1
#| fig-height: 12
tar_read(viz_sankey_act_resp_stage1_ggplot_fig)
```

### Stage 2 — cited-literature overlap (no embeddings)

```{r}
#| label: fig-sankey-act-resp-stage2
#| fig-height: 12
tar_read(viz_sankey_act_resp_stage2_ggplot_fig)
```

### Stage 3 — cited-literature embedding similarity

```{r}
#| label: fig-sankey-act-resp-stage3
#| fig-height: 12
tar_read(viz_sankey_act_resp_stage3_ggplot_fig)
```

## TCA Approaches ↔ TCA Actions Ch5

Directed **TCA Approaches → TCA Action**: for each Approach, its top 3
most-similar Actions under each signal stage.

### Stage 1 — definition-text similarity (embeddings)

```{r}
#| label: fig-sankey-appr-act-stage1
#| fig-height: 12
tar_read(viz_sankey_appr_act_stage1_ggplot_fig)
```

### Stage 2 — cited-literature overlap (no embeddings)

```{r}
#| label: fig-sankey-appr-act-stage2
#| fig-height: 12
tar_read(viz_sankey_appr_act_stage2_ggplot_fig)
```

### Stage 3 — cited-literature embedding similarity

```{r}
#| label: fig-sankey-appr-act-stage3
#| fig-height: 12
tar_read(viz_sankey_appr_act_stage3_ggplot_fig)
```
