---
title: "NXS TCS Article — Citation Method Comparison"
subtitle: "Regex vs LLM (OpenRouter) identification of in-text citations in the definitions"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(dplyr)
library(knitr)
```

The cited literature is discovered by **extracting in-text citations from the
concept definitions** and resolving each to a work in the matching assessment's
reference library. Two extraction methods run as parallel pipeline targets:

- **regex** — a deterministic parser (`R/citation_regex.R`) matching
  parenthetical and narrative `(Author, year)` forms.
- **llm** — an OpenRouter chat model (`R/extract_definition_citations_llm.R`)
  reading each definition and returning its citations as JSON.

Both feed the *same* two-pass resolver, so the resolved tables are directly
comparable. There is no hand-labelled ground truth, so "better" is judged by
three complementary signals:

1. **Volume** — how many citations each finds (a recall proxy).
2. **Agreement** — citations found by *both* are high-confidence; method-only
   citations are where the methods disagree.
3. **Resolution rate** — the fraction of what each finds that resolves to a
   real work in the assessment library (a precision proxy: a hallucinated or
   mis-parsed citation is unlikely to resolve).

# Headline metrics

```{r}
#| label: tbl-summary
tar_read(viz_citation_summary_tbl) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = paste(
      "Per method: total citation tokens extracted, how many resolved",
      "(strict / loose), resolution %, and distinct works cited. Higher",
      "resolution % suggests cleaner extraction; more distinct works = wider",
      "cited-literature coverage."
    )
  )
```

Read this together: the method with **more citations *and* a comparable or
higher resolution %** is finding more real citations rather than noise. If one
method extracts many more but at a much lower resolution %, the extras are
likely spurious (mis-parses or hallucinations that match nothing).

# Agreement per keyset

```{r}
#| label: fig-overlap
#| fig-cap: "Distinct (author, year) citations per keyset, split by whether they were found by both methods, regex only, or LLM only. A large 'both' block = the methods largely agree; large method-only blocks locate systematic differences (e.g. narrative citations, institutional authors)."
tar_read(viz_citation_overlap_fig)
```

```{r}
#| label: tbl-overlap
tar_read(viz_citation_overlap_data) |>
  tidyr::pivot_wider(names_from = agreement, values_from = n,
                     values_fill = 0L) |>
  dplyr::mutate(
    total = rowSums(dplyr::across(dplyr::where(is.numeric))),
    `agreement %` = round(100 * `both` / total, 1)
  ) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = "Citation agreement counts per keyset (both / regex only / llm only)."
  )
```

# Citations to review: disagreements & unmatched

Every citation worth a look — either the two methods **disagree** (found by only
one), or it stayed **unmatched** (resolved by neither method that found it).
Rows found by *both* methods *and* resolved are omitted (nothing to check).
Columns: `found_by` (regex only / llm only / both), `resolved`, `pass`
(strict / loose / unmatched), and the **actual citation text**. A method-only
citation that doesn't resolve is the least trustworthy — a mis-parse (regex) or
an over-eager pickup (LLM); one that does resolve is a genuine citation the
other method missed. `both` + unmatched = a real citation neither library
contains (often institutional authors, e.g. IUCN / UNFCCC) — the manual
verification set. Searchable/sortable:

```{css}
/*| echo: false */
/* Shrink the disagreement DT (the only DT in this report): header, body,
   filter row and controls. */
.dataTables_wrapper,
table.dataTable,
table.dataTable th,
table.dataTable td,
.dataTables_wrapper .dataTables_filter input {
  font-size: 10px;
  line-height: 1.2;
}
```

```{r}
#| label: tbl-review
d <- tar_read(viz_citation_review)
DT::datatable(
  d,
  rownames = FALSE,
  filter = "top",
  class = "compact stripe hover",
  options = list(pageLength = 25, order = list(list(0, "asc"))),
  caption = sprintf(
    "%d citations to review (disagreements + unmatched): found_by, resolved, actual text.",
    nrow(d)
  )
)
```

# Reading the verdict

- If **both** dominates and resolution % is similar, the cheaper, deterministic
  **regex** method is sufficient — keep `citations.active: regex`.
- If the **LLM** finds materially more citations that *also* resolve at a
  similar or higher rate (especially narrative citations regex misses),
  consider switching `citations.active: llm` in `input/config.yaml`.
- Method-only citations that **fail to resolve** are the least trustworthy —
  weigh a method down if most of its extra finds are unmatched.

# Appendix — how each method identifies citations

```{r}
#| label: load-methods
#| include: false
# Source the extractor code so the exact prompt / patterns are shown verbatim
# (single source of truth — no risk of the report drifting from the code).
source("R/citation_regex.R")
source("R/extract_definition_citations_llm.R")
llmc <- yaml::read_yaml("input/config.yaml")$citations$llm
```

## LLM prompt (OpenRouter)

Each definition is sent to the model with the system prompt below (verbatim
from `R/extract_definition_citations_llm.R`); the user message is just
`TEXT:\n<definition>`, and the reply is parsed as JSON.

```{r}
#| label: llm-prompt
#| echo: false
#| results: asis
cat(sprintf(
  "**Model:** `%s` — temperature %s, timeout %ss, %s parallel workers.\n\n",
  llmc$model, llmc$temperature, llmc$timeout, llmc$workers
))
cat("```\n", .LLM_CITE_PROMPT, "\n```\n", sep = "")
```

## Regex rules

The regex extractor (`R/citation_regex.R`) has no prompt; it scans each
definition for:

- **Parenthetical** groups `( … )` that contain a year, split on `;` into
  individual citations (e.g. `(Kabeer, 1999; Manlosa, 2022)` → two);
- **Narrative** citations — `Surname (year)`, `Surname et al. (year)`,
  `A & B (year)`.

In both cases the first-author surname is normalised (lower-cased, diacritics
folded, initials / "et al." dropped) into the `author_key` used for resolution.
The year is matched with:

```{r}
#| label: regex-year
#| echo: false
#| results: asis
cat("```\n", .CITE_YEAR, "\n```\n", sep = "")
```
