---
title: "NXS TCS Article — Reports"
subtitle: "IPBES TCA & Nexus assessment corpus: embeddings, keypaper similarity, topics"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(targets)
# Depend on the two reports (tar_read) so this index is built *after* them and
# its links resolve. report_embeddings / report_analysis return the exact
# rendered .html paths (used for their links). Topic modelling is disabled for
# now, so there is no topic-report dependency.
emb_path <- tar_read(report_embeddings)
ana_path <- tar_read(report_analysis)
cmp_path <- tar_read(report_citation_comparison)
# Pipeline diagrams — pre-rendered SVGs (mmd_figs, via mermaid-cli) embedded
# below. tar_read makes editing a .mmd (→ mmd_figs) re-render this index.
mmd_figs <- tar_read(mmd_figs)
pipeline_svg <- grep("/pipeline\\.svg$", mmd_figs, value = TRUE)[1]
sequence_svg <- grep("/sequence\\.svg$", mmd_figs, value = TRUE)[1]
enc <- function(x) gsub(" ", "%20", x) # href-encode spaces in the filenames
emb_html <- enc(basename(emb_path))
ana_html <- enc(basename(ana_path))
cmp_html <- enc(basename(cmp_path))
```

This is the landing page for the **NXS TCS Article** analysis. The corpus is
built from two curated IPBES assessment Zotero groups — **TCA** and **NXS** —
matched to OpenAlex works and extracted from a local snapshot. Each work is
embedded with **SPECTER2** (title, abstract, and title+abstract variants) and
scored by cosine similarity against a curated set of **TCA/Nexus keypaper
concept definitions**. Inline citations in the definitions are extracted and
resolved to a **cited corpus**, and the key-sets are linked to one another
(notably **TCA Action ↔ Nexus Response Option**) via three similarity stages.

The analysis is split across three reports, described below. Each opens in the
`output/reports/` folder alongside this page; a link resolves once that report
has been rendered (`targets::tar_make()`).

```{r}
#| label: links
#| output: asis
cat(sprintf(
  '
### [Embedding Report](%s)

Embedding **quality control**: corpus coverage (title / abstract presence and
counts), language mix, input text length and silent truncation at the 512-token
limit, work-type counts, title vs abstract variant agreement, and embedding-norm
sanity checks. Read this to judge whether the embeddings are trustworthy before
interpreting the analysis.

### [Chapter Analysis Report](%s)

The main analysis: **how close the papers in each assessment / chapter are to
the keypaper concepts**, plus **keyset linkage** — how the concept definitions
link to one another (notably TCA Action ↔ Nexus Response Option) across three
stages (definition-embedding similarity, cited-literature overlap, and
cited-literature embedding similarity) with Sankey diagrams. Also overall
chapter alignment (median similarity with bootstrap 95%% CIs), Cliff’s
δ, the chapter × keypaper heatmap, an interactive per-keypaper
top-matches browser, corpus-wide keypaper-relevance, and keypaper-set
diagnostics.

_Topic modelling (BERTopic) is disabled in this configuration._

### [Citation Method Comparison Report](%s)

Compares the **two citation-identification methods** — the deterministic regex
parser and the LLM (OpenRouter) parser — that extract the in-text citations
from the definitions: extraction volume, agreement per keyset, and resolution
rate (a precision proxy) to judge which performs better.

',
  emb_html,
  ana_html,
  cmp_html
))
```

# Pipeline

The `targets` pipeline that produces these reports — from the two IPBES Zotero
groups through embeddings, keypaper scoring, citation resolution and keyset
linkage. (Source: `input/mmd/pipeline.mmd`.)

```{r}
#| label: fig-pipeline
#| fig-cap: "The `targets` pipeline (rendered from `input/mmd/pipeline.mmd`)."
knitr::include_graphics(pipeline_svg)
```

## Run sequence

The same run as a sequence diagram (who talks to whom during `tar_make()`).
(Source: `input/mmd/sequence.mmd`.)

```{r}
#| label: fig-sequence
#| fig-cap: "The same run as a sequence diagram (rendered from `input/mmd/sequence.mmd`)."
knitr::include_graphics(sequence_svg)
```
