---
title: "NXS TCS Article — Reports"
subtitle: "IPBES TCA & Nexus assessment corpus: embeddings, keypaper similarity, topics"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: false
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(targets)
# Depend on the three reports (tar_read) so this index is built *after* them
# and its links resolve. report_embeddings / report_analysis return the exact
# rendered .html paths (used for their links); report_topic_modelling is the
# active topic report — a dependency token that guarantees at least the active
# topic report exists (the glob below then also lists any archived ones).
# NOTE: this couples the index to report_topic_modelling, which needs a
# BERTopic pod run — building the index will trigger that pipeline.
emb_path <- tar_read(report_embeddings)
ana_path <- tar_read(report_analysis)
invisible(tar_read(report_topic_modelling))
# Pipeline diagrams — pre-rendered SVGs (mmd_figs, via mermaid-cli) embedded
# below. tar_read makes editing a .mmd (→ mmd_figs) re-render this index.
mmd_figs <- tar_read(mmd_figs)
pipeline_svg <- grep("/pipeline\\.svg$", mmd_figs, value = TRUE)[1]
sequence_svg <- grep("/sequence\\.svg$", mmd_figs, value = TRUE)[1]
cfg <- yaml::read_yaml(tar_read(config_file))
enc <- function(x) gsub(" ", "%20", x) # href-encode spaces in the filenames
emb_html <- enc(basename(emb_path))
ana_html <- enc(basename(ana_path))

# ALL rendered topic-modelling reports (one per bertopic config, archived by
# render_report_topic_modelling), discovered in output/reports/ rather than
# only the active_for_viz one. Re-render this index to pick up new ones.
report_dir <- "output/reports"
topic_pat <- "^NXS TCS Article Topic Modelling Report - (.*)\\.html$"
topic_files <- if (dir.exists(report_dir)) {
  sort(list.files(report_dir, pattern = topic_pat))
} else {
  character(0)
}
topic_cfg <- sub(topic_pat, "\\1", topic_files) # bertopic config name each
active_viz <- cfg$bertopic$active_for_viz
```

This is the landing page for the **NXS TCS Article** analysis. The corpus is
built from two curated IPBES assessment Zotero groups — **TCA** and **NXS** —
matched to OpenAlex works and extracted from a local snapshot. Each work is
embedded with **SPECTER2** (title, abstract, and title+abstract variants) and
scored by cosine similarity against a curated set of **TCA/Nexus keypaper
concept definitions**. The corpus is then clustered with **BERTopic**.

The analysis is split across three reports, described below. Each opens in the
`output/reports/` folder alongside this page; a link resolves once that report
has been rendered (`targets::tar_make()`).

```{r}
#| label: links
#| output: asis
cat(sprintf(
  '
### [Embedding Report](%s)

Embedding **quality control**: corpus coverage (title / abstract presence and
counts), language mix, input text length and silent truncation at the 512-token
limit, work-type counts, title vs abstract variant agreement, and embedding-norm
sanity checks. Read this to judge whether the embeddings are trustworthy before
interpreting the analysis.

### [Chapter Analysis Report](%s)

The main analysis: **how close the papers in each assessment / chapter are to
the keypaper concepts.** Overall chapter alignment (median similarity with
bootstrap 95%% CIs), which chapters are *really* different (Cliff’s
δ), the chapter × keypaper heatmap, and an interactive per-keypaper
top-matches browser with matches stacked by assessment / chapter. Also includes
corpus-wide keypaper-relevance (score distributions, score vs year / citations /
work type) and keypaper-set diagnostics (coherence, per-keypaper distribution).

### Topic Modelling Reports

**BERTopic** clustering of the corpus (UMAP + HDBSCAN) and how the keypaper
concepts distribute across the discovered topics. One report per BERTopic
config; the config name is shown in front of each link.

',
  emb_html,
  ana_html
))

if (length(topic_files)) {
  for (i in seq_along(topic_files)) {
    tag <- if (identical(topic_cfg[i], active_viz)) {
      " — *active (feeds the viz layer)*"
    } else {
      ""
    }
    cat(sprintf(
      "- **%s** — [Topic Modelling Report](%s)%s\n",
      topic_cfg[i],
      enc(topic_files[i]),
      tag
    ))
  }
} else {
  cat(
    "_No topic-modelling reports have been rendered yet — run the topic",
    "pipeline, then re-render this index._\n"
  )
}
```

# Pipeline

The `targets` pipeline that produces these reports — from the two IPBES Zotero
groups through embeddings, keypaper scoring and BERTopic. (Source:
`input/mmd/pipeline.mmd`.)

```{r}
#| label: fig-pipeline
#| fig-cap: "The `targets` pipeline (rendered from `input/mmd/pipeline.mmd`)."
knitr::include_graphics(pipeline_svg)
```

## Run sequence

The same run as a sequence diagram (who talks to whom during `tar_make()`).
(Source: `input/mmd/sequence.mmd`.)

```{r}
#| label: fig-sequence
#| fig-cap: "The same run as a sequence diagram (rendered from `input/mmd/sequence.mmd`)."
knitr::include_graphics(sequence_svg)
```
