---
title: "NXS TCS Article — Embedding Report"
subtitle: "SPECTER2 embeddings, keypaper-similarity scoring"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    grid:
        sidebar-width: 0px
        body-width: 4000px
        margin-width: 200px
        gutter-width: 1.5rem      
execute:
  message: false
  warning: false
  cache: false
---

```{r}
#| label: setup
library(targets)
library(htmltools)
library(knitr)
```


# Embeddings Overview
The corpus contains the following missing titles and abstracts

```{r}
#| label: tbl-missing-titles
ds <- arrow::open_dataset(tar_read(corpus))
n_total <- nrow(ds)
n_na <- ds |> dplyr::filter(is.na(title)) |> nrow()
n_empty <- ds |> dplyr::filter(!is.na(title), title == "") |> nrow()
n_present <- n_total - n_na - n_empty

tibble::tibble(
  state = c("title present", "title empty (nchar=0)", "title NA", "total"),
  n = c(n_present, n_empty, n_na, n_total)
) |>
  dplyr::mutate(`%` = 100 * n / n_total) |>
  knitr::kable(
    digits = 3,
    format.args = list(big.mark = ",", nsmall = 3),
    align = c("l", "r", "r"),
    caption = "Title presence in the OpenAlex corpus extract. Works in NA or empty rows are dropped from the title-variant embedding by preprocessor_title()."
  )
```

```{r}
#| label: tbl-missing-abstracts
ds <- arrow::open_dataset(tar_read(corpus))
n_total <- nrow(ds)
n_na <- ds |> dplyr::filter(is.na(abstract)) |> nrow()
n_empty <- ds |> dplyr::filter(!is.na(abstract), abstract == "") |> nrow()
n_present <- n_total - n_na - n_empty

tibble::tibble(
  state = c(
    "abstract present",
    "abstract empty (nchar=0)",
    "abstract NA",
    "total"
  ),
  n = c(n_present, n_empty, n_na, n_total)
) |>
  dplyr::mutate(`%` = 100 * n / n_total) |>
  knitr::kable(
    digits = 3,
    format.args = list(big.mark = ",", nsmall = 3),
    align = c("l", "r", "r"),
    caption = "Abstract presence in the OpenAlex corpus extract. Works without abstracts are dropped from the abstract and title_abstract variants."
  )
```

The following number of embeddings were determined.

```{r}
#| label: metadata
# Embeddings root holds multiple config=<name>/ partitions

emb_meta <- dirname(dirname(dirname(tar_read(emb_corpus_title)))) |>
  arrow::open_dataset() |>
  dplyr::count(config, source, variant) |>
  dplyr::collect() |>
  tidyr::pivot_wider(names_from = variant, values_from = n, values_fill = 0L) |>
  dplyr::relocate(
    dplyr::any_of(c("title", "abstract", "title_abstract")),
    .after = source
  ) |>
  dplyr::arrange(config, source)


# Total works available per source (before embedding)
n_corpus <- nrow(arrow::open_dataset(tar_read(corpus)))
n_keypaper <- nrow(arrow::open_dataset(tar_read(key_works)))

emb_meta <- emb_meta |>
  dplyr::mutate(
    corpus_size = dplyr::if_else(source == "corpus", n_corpus, n_keypaper)
  ) |>
  dplyr::arrange(config, source) |>
  dplyr::mutate(
    corpus_size = dplyr::if_else(source == "corpus", n_corpus, n_keypaper)
  ) |>
  dplyr::relocate(corpus_size, .after = source)

# Map config name → model_id for the caption
cfg <- yaml::read_yaml("input/config.yaml")
active_cfg <- cfg$embeddings$active
model_ids <- vapply(
  cfg$embeddings$configs,
  function(c) c$model %||% NA_character_,
  character(1)
)

knitr::kable(
  emb_meta,
  format.args = list(big.mark = ","),
  caption = sprintf(
    "Embedded works per config × source × variant. Active config: %s (model_id = %s).",
    active_cfg,
    model_ids[[active_cfg]]
  )
)
```

## Language distribution

SPECTER2 is a primarily English model; non-English works can still
embed but their similarity scores are less reliable. Useful gut-check
on how much of the corpus the embedding is well-suited for.

```{r}
#| label: fig-language
#| fig-cap: "Works per OpenAlex `language`, log x-axis. Blue bars = ≥ 0.5% of the corpus; grey = long tail. SPECTER2 was trained primarily on English; treat similarity scores for low-resource languages with caution."
tar_read(viz_language_fig)
```

## Input text length per variant

How long is the text actually sent to TEI? `title_abstract` is capped on
the title side at `title_cap_combined` chars (config), then joined with
the abstract via `[SEP]`. Works far above ~2000 chars may hit the model's
512-token context window and get silently truncated server-side.

```{r}
#| label: fig-text-length
#| fig-cap: "Distribution of input text length per variant (characters, log scale). Title is by definition short; abstract dominates the right tail; title_abstract is title (capped) + sep + abstract."
tar_read(viz_text_length_fig)
```

### Silent truncation by TEI

SPECTER2 truncates input above ~512 tokens (~2,000 characters)
**without warning**. Embeddings for truncated works are based only on
the first 2k characters of text — methodologically worth surfacing.

```{r}
#| label: tbl-truncation
tar_read(viz_truncation_stats) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = "Number of works exceeding length thresholds per variant. The `> 2048` column approximates how many were silently truncated by TEI."
  )
```


# Work-type breakdown

The OpenAlex corpus contains 19 work types — overwhelmingly `article`,
plus reviews, book chapters, preprints and a long tail of small types
(letters, errata, paratext, …). Counts per type set the baseline; the
score-by-type breakdown lives in the Chapter Analysis report.

```{r}
#| label: fig-type-counts
#| fig-cap: "Number of works per OpenAlex `type`, log x-axis. Blue bars = types representing ≥ 0.5% of the corpus; grey = long-tail types."
tar_read(viz_type_count_fig)
```

# Variant agreement

If `title` and `abstract` agree on a work's relevance, the title is a reliable
proxy for the abstract. Off-diagonal works are interesting candidates for
manual inspection: either the title misrepresents the paper or the abstract is
poor quality.

```{r}
#| label: fig-variant-agree
#| fig-cap: "Per-work max similarity to keypapers, plotted pair-wise across the three variants. Each corpus work contributes one point per pair (coloured). The dashed diagonal is perfect agreement; points off it are works the two variants disagree about. Click or box/lasso-select to highlight a work's points across all pairs."
tar_read(viz_agree_fig)
```

# Sanity check: embedding L2 norms

SPECTER2 should return roughly normalised vectors (norm ≈ 1). A tight
peak near 1 means the model behaved as expected; outliers or zero-norm
vectors would indicate problems (empty input, model failures). Sampled
to 50k per (source, variant) — sufficient for a histogram.

```{r}
#| label: fig-emb-norm
#| fig-cap: "Distribution of L2 norms of embedding vectors per (source × variant), sampled to 50,000 rows per leaf, seed = 13. Expect a tight peak — outliers or wide spread suggest input or model issues."
tar_read(viz_emb_norm_fig)
```

# Provenance

Everything needed to reproduce this report.

```{r}
#| label: tbl-provenance
cfg <- yaml::read_yaml("input/config.yaml")
active_cfg <- cfg$embeddings$active
emb_cfg <- cfg$embeddings$configs[[active_cfg]]

# Embedding dimension from one parquet file
emb_dim <- {
  one_file <- list.files(
    dirname(tar_read(emb_corpus_title)),
    pattern = "\\.parquet$",
    full.names = TRUE,
    recursive = TRUE
  )[[1]]
  v_cols <- grep(
    "^V[0-9]+$",
    names(arrow::open_dataset(one_file)),
    value = TRUE
  )
  length(v_cols)
}

git_sha <- tryCatch(
  trimws(system("git rev-parse --short HEAD", intern = TRUE)),
  error = function(e) "unknown"
)

tibble::tibble(
  field = c(
    "active embedding config",
    "model id",
    "embedding dimension",
    "title cap (chars, combined variant)",
    "separator token",
    "TEI host",
    "report rendered",
    "git commit"
  ),
  value = c(
    active_cfg,
    emb_cfg$model %||% "—",
    as.character(emb_dim),
    as.character(emb_cfg$title_cap_combined %||% "—"),
    emb_cfg$sep_token %||% "—",
    paste0(emb_cfg$scheme %||% "http", "://", emb_cfg$host),
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"),
    git_sha
  )
) |>
  knitr::kable(
    caption = "Run provenance — paste into the methods section verbatim."
  )
```
