---
title: "NXS TCS Article — Chapter Analysis"
subtitle: "How do the keypaper concepts link across the TCA/Nexus key-sets?"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    grid:
        sidebar-width: 0px
        body-width: 4000px
        margin-width: 200px
        gutter-width: 1.5rem
execute:
  message: false
  warning: false
  cache: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(htmltools)
library(knitr)
```

This report examines **how the keypaper concepts (the TCA/Nexus definitions)
link to one another across the four key-sets** — especially the directed link
**TCA Approaches → TCA Action** — and how coherent the keypaper set itself is.
It builds toward the combined **Approach × Action** heatmap at the end.

The heavier corpus-wide analyses (chapter alignment, score distributions,
work-type and citation breakdowns) and the cited-literature-embedding "Stage 3"
signal have been removed: they depend on the full-corpus embeddings, which are
no longer computed in this pruned pipeline.

# Keypaper set

## Keypaper coherence

How tight is the keypaper set itself? Pairwise cosine similarity between the
keypapers (title_abstract variant), reordered by average-linkage hierarchical
clustering. A coherent set shows clear high-similarity blocks on the diagonal.

```{r}
#| label: fig-keypaper-self-sim
#| fig-cap: "Pairwise cosine similarity between keypapers (title_abstract variant). Axis order is from hierarchical clustering (1 − cosine, average linkage). Block structure on the diagonal indicates topical sub-groups within the keypaper set."
tar_read(viz_keypaper_self_sim_fig)
```

# Keyset linkage

How do the concept definitions link *to one another* across the four key-sets
(the worksheet groups: `TCA_Annex_3_3`, `TCA_Approaches_3_2`,
`TCA_Actions_Ch5`, `Nexus_Response_Options`)? Two directed links are of
particular interest — **TCA Action → Nexus Response Option** and
**TCA Approaches → TCA Action** — each shown below as a set of Sankeys. Two
complementary similarity signals drive them:

- **Stage 1 — definition-text similarity (embeddings):** cosine similarity
  between the keypaper definition embeddings (title_abstract, SPECTER2). Two
  definitions link if their *wording/concepts* are close.
- **Stage 2 — cited-literature overlap (no embeddings):** two definitions link
  if they *cite the same works* — Jaccard overlap (Jaccard 1912) of the
  resolved cited-work sets (only pairs sharing ≥1 work).

Jaccard, P. (1912). The distribution of the flora in the alpine zone.
*New Phytologist*, 11(2), 37–50.
<https://doi.org/10.1111/j.1469-8137.1912.tb05611.x> (the "coefficient of
community", defined on p. 39: species common to both / total species in both).

## Keyset × keyset overview

Each Sankey below isolates one source → target keyset pair. These matrices give
the whole picture: the mean similarity between *every* pair of key-sets, per
signal stage (read a row against the columns).

```{r}
#| label: fig-keyset-matrix-stage1
#| fig-cap: "Stage 1 — keyset × keyset mean definition-embedding cosine similarity."
tar_read(viz_keyset_matrix_stage1_fig)
```

```{r}
#| label: fig-keyset-matrix-stage2
#| fig-cap: "Stage 2 — keyset × keyset mean Jaccard overlap of cited-work sets."
tar_read(viz_keyset_matrix_stage2_fig)
```

## TCA Approaches ↔ TCA Actions Ch5

Directed **TCA Approaches → TCA Action**: for each Approach, its top
most-similar Actions under each signal stage.

The Sankeys above only show each Approach's top matches; the heatmaps below
show every Approach × Action pair for both signals. Black borders mark
cells scoring above the 95th percentile of that measure's corpus-wide
pairwise distribution — i.e. a stronger link than typical, and unlikely to be
chance; unbordered cells are indistinguishable from background.

### Stage 1 — definition-text similarity (embeddings)

```{r}
#| label: fig-sankey-appr-act-stage1
#| fig-height: 12
tar_read(viz_sankey_appr_act_stage1_ggplot_fig)
```

```{r}
#| label: fig-pair-heatmap-appr-act-stage1
#| fig-cap: "Approach × Action, Stage 1 (definition-text cosine similarity). Black border = above the 95th percentile of all corpus-wide Stage 1 pairwise values."
#| fig-height: 10
tar_read(viz_pair_heatmap_appr_act_stage1_fig)
```

### Stage 2 — cited-literature overlap (no embeddings)

```{r}
#| label: fig-sankey-appr-act-stage2
#| fig-height: 12
tar_read(viz_sankey_appr_act_stage2_ggplot_fig)
```

```{r}
#| label: fig-pair-heatmap-appr-act-stage2
#| fig-cap: "Approach × Action, Stage 2 (cited-literature Jaccard overlap). Black border = above the 95th percentile of all corpus-wide Stage 2 pairwise values (background includes the many zero-overlap pairs)."
#| fig-height: 10
tar_read(viz_pair_heatmap_appr_act_stage2_fig)
```

### Combined (Stage 1 + Stage 2, equal weight)

Each cell blends Stage 1 and Stage 2 as equal-weight indicators — each stage's
value expressed as a percentile rank against its own corpus-wide background,
then averaged, so neither measure's raw scale dominates.

```{r}
#| label: fig-pair-heatmap-appr-act-combined
#| fig-height: 10
tar_read(viz_pair_heatmap_appr_act_combined_fig)
```

*Approach × Action, combined Stage 1 + Stage 2 (equal-weight mean
percentile). **Center number** (large) = the combined score; **bottom-left**
(small) = Stage 1 definition-text cosine similarity; **bottom-right** (small)
= Stage 2 cited-literature Jaccard overlap. **Bold** marks a value that is
itself above its own measure's 95th percentile (a real link); the center
number is bold whenever either component is. Border style (solid = both
stages agree, dashed = one stage only, no border = neither) is keyed in the
figure's own "Individual-stage link" legend.

**Fill colour and border answer two different questions, and they need not
agree.** The fill tracks the **center number** — *how unusual* this pair is —
on ColorBrewer's colour-blind-safe RdBu palette. Because the combined score
averages two percentile ranks taken against the background of all keypaper
pairs, **white sits at 0.5, meaning "as connected as a typical pair of
concept definitions"**: red is more connected than typical, blue less. Note
this is a *relative* measure — a Jaccard overlap of 0.03 is tiny in absolute
terms but lands high because most concept pairs share no citations at all.

The border answers instead whether an **individual stage** clears its own
95th percentile, i.e. whether the link is strong enough to be treated as
real, and the **small bottom numbers** are bold on the same test. A cell can
therefore be distinctly red yet carry no border — sitting, say, at the 93rd
percentile on both stages: high, but short of the significance bar. The
colour ranks; the border and bold marks alone certify. The
"Jaccard overlap" is Jaccard's coefficient of
community (Jaccard 1912, p. 39; species common to both / total species in
both) — the term "Jaccard index" is a later name, not used in the original
paper.*
