---
title: "NXS TCS Article — Chapter Analysis"
subtitle: "How close are the assessment chapters to the keypaper concepts?"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    grid:
        sidebar-width: 0px
        body-width: 4000px
        margin-width: 200px
        gutter-width: 1.5rem
execute:
  message: false
  warning: false
  cache: false
---

```{r}
#| label: setup
#| include: false
library(targets)
```

This report measures **how similar the papers in each assessment chapter are to
the keypaper concepts** (the TCA/Nexus definitions). It is driven by the
combined per-chapter scores (`scores_combined`): each corpus work scored from
its `title_abstract` embedding, or its `title` embedding where it has no
abstract, against the keypaper set — one row per `(work, assessment, chapter)`.

A work tagged with several Zotero chapters is counted **once per chapter**
(honest for "which chapters supply the matches"). Chapter labels are shown as
tagged, except the malformed NXS tag `Chapter(s) 5`, relabelled to `Chapter 5`;
the `Chapter 5.0…5.6` sub-chapters are kept separate.

# Overall: which chapters are closest to the keypapers?

Median (with inter-quartile bar) of each work's maximum cosine similarity to
its nearest keypaper, per assessment/chapter. Higher = the chapter's papers sit
closer to the keypaper concepts overall.

```{r}
#| label: fig-chapter-alignment
#| fig-cap: "Chapter alignment to the keypaper set: median (bar) and bootstrap 95% CI of the median (error bar) of per-work max cosine similarity to the nearest keypaper, per assessment/chapter, sorted ascending. Chapters whose CIs do not overlap are really different; small-n chapters get wide CIs. n = works × chapter memberships in that chapter."
#| fig-height: 8
tar_read(viz_chapter_alignment_fig)
```

## Which chapters are *really* different? (Cliff's δ)

Raw cosine gaps are hard to read on the compressed similarity scale, and with
thousands of works trivial gaps look "significant". Cliff's δ is an effect
size: for a pair of chapters it is P(a row-chapter work scores higher) −
P(it scores lower). **0 ≈ indistinguishable**; |δ| ≈ 0.15 / 0.33 / 0.47 is a
small / medium / large separation. Read a row against the columns: red = the
row chapter's papers tend to score *higher* than the column's, blue = lower.

```{r}
#| label: fig-chapter-cliffs-delta
#| fig-cap: "Pairwise Cliff's δ between assessment/chapters on per-work max keypaper similarity. Rows/columns ordered by median alignment (highest first). δ near 0 (white) = the two chapters' similarity distributions are indistinguishable."
#| fig-height: 9
tar_read(viz_chapter_cliffs_delta_fig)
```

# In detail: chapter × keypaper

Median similarity of each assessment/chapter to each **individual** keypaper.
Bright cells = the chapter's papers are close to that concept; dark rows =
concepts the corpus has little near (orphans); a chapter's brightest cells are
the concepts it most aligns with. Rows (keypapers) and columns (chapters) are
ordered by overall median.

```{r}
#| label: fig-chapter-keypaper-heatmap
#| fig-cap: "Median cosine similarity per (assessment/chapter, keypaper). Rows = keypaper concepts, columns = assessment/chapter."
#| fig-height: 12
tar_read(viz_chapter_keypaper_heatmap_fig)
```

# Top matches per keypaper, by chapter

Pick a keypaper concept to browse its highest-scoring corpus matches. The
histogram stacks the matches' similarity by **assessment/chapter**, so you can
see which chapters supply a concept's closest papers. **Toggle chapters in the
legend** (click to show/hide, double-click to isolate); `No Chapter` works are
**hidden by default** — click their legend entry to include them. The table
below carries `assessment` and `chapter` columns (filterable), and the CSV
button exports the currently filtered rows. Links go to the DOI when available,
otherwise the OpenAlex page.

```{r}
#| label: tbl-top-matches-combined
tar_read(tbl_top_matches_combined)
```
