---
title: "NXS TCS Article — Topic Modelling Report"
subtitle: "BERTopic clustering of the TCAC 2.0 corpus (UMAP + HDBSCAN)"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
execute:
  message: false
  warning: false
  cache: false
---

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.11389482.svg)](https://doi.org/10.5281/zenodo.11389482)
[![GitHub release](https://img.shields.io/github/release/IPBES-Data/IPBES_TCA_Corpus.svg)](https://github.com/IPBES-Data/IPBES_TCA_Corpus/releases/latest)
[![License: CC BY 4.0](https://img.shields.io/badge/License-CC%20BY%204.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

Part of the Data Management Report [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10251349.svg)](https://doi.org/10.5281/zenodo.10251349)

# Disclaimer

# Working Title

TCAC 2.0 — Transformative Change Assessment Corpus 2.0

## Code repo

[Github repository](https://github.com/IPBES-Data/TCAC-2.0)

# Introduction

This report documents the topic-modelling layer of TCAC 2.0. SPECTER2
embeddings of the corpus (produced and assessed in the
[Embedding Report](TCAC%202.0%20Embedding%20Report.html)) are reduced
to a low-dimensional space with UMAP and clustered with HDBSCAN, using
the [BERTopic](https://maartengr.github.io/BERTopic/) framework. Each
work in the corpus and each keypaper is assigned a `topic_id`. Topics
are summarised by per-topic c-TF-IDF vocabulary on the title +
abstract text.

The full-corpus fit (~5.77 M works) runs on a RunPod GPU pod via cuml
(`topics_runpod`). Outputs are written hive-partitioned under
`output/NXS_TCA_corpus/topics/config=<emb>/bertopic=<run>/variant=<v>/`.

```{r}
#| label: setup
library(targets)
library(arrow)
library(dplyr)
library(ggplot2)
library(htmltools)
library(knitr)
```

# Method

## Pipeline summary

The pipeline is orchestrated by `targets`. Stages relevant to this
report:

1. **Embeddings** — SPECTER2 (768-d) on title, abstract and
   title+abstract variants. See the
   [Embedding Report](TCAC%202.0%20Embedding%20Report.html).
2. **UMAP** — reduce the primary variant (`title_abstract`) to
   `n_components = 5` dimensions for clustering, plus a 2-D projection
   for visualisation.
3. **HDBSCAN** — density-based clustering on the 5-D UMAP space.
4. **c-TF-IDF** — per-topic term scoring on the title+abstract text,
   producing the topic vocabulary.
5. **Fallback projection** — works without abstracts (and the keypapers
   without abstracts) are projected via the title variant onto the same
   UMAP+HDBSCAN fit and tagged `topic_source = "fallback"`.

## Parameters used

The BERTopic fit is identified by three things: the **embedding
config** it consumes (which embedding model + which run produced the
vectors), the **variant** of those embeddings used as the primary
input, and the **BERTopic config** itself (UMAP + HDBSCAN +
vectorizer knobs). All three are read from `input/config.yaml`.

```{r}
#| label: tbl-bertopic-cfg
y <- yaml::read_yaml("input/config.yaml")
active_emb_cfg <- y$embeddings$active
active_btopic <- y$bertopic$active_runpod
cfg <- tar_read(bertopic_runpod_cfg)

# Show only the knobs that affect the fit (skip transport/ssh details).
knobs <- c(
  "method",
  "primary_variant",
  "fallback_variant",
  "center_embeddings",
  "umap_n_components",
  "umap_n_neighbors",
  "umap_min_dist",
  "umap_metric",
  "hdbscan_min_cluster_size",
  "hdbscan_min_samples",
  "top_n_words",
  "vectorizer_min_df",
  "vectorizer_max_df",
  "vectorizer_max_features",
  "vectorizer_ngram",
  "keypaper_threshold",
  "random_seed"
)
have <- intersect(knobs, names(cfg))

tibble::tibble(
  parameter = c("embedding config", "bertopic config", have),
  value = c(
    active_emb_cfg,
    active_btopic,
    vapply(
      have,
      function(k) {
        v <- cfg[[k]]
        if (is.null(v)) "—" else paste(v, collapse = ", ")
      },
      character(1)
    )
  )
) |>
  knitr::kable(
    caption = sprintf(
      "Active embedding config: %s · BERTopic config: %s · Primary variant: %s · Fallback variant: %s.",
      active_emb_cfg,
      active_btopic,
      cfg$primary_variant %||% "title_abstract",
      cfg$fallback_variant %||% "title"
    )
  )
```

# Results

## Topic count and corpus coverage

```{r}
#| label: tbl-topic-summary
# tar_read(topics_runpod) returns the path to topic_info.parquet
# inside the hive-partitioned leaf
# output/NXS_TCA_corpus/topics/config=<emb>/bertopic=<run>/variant=<v>/.
# Parse the three hive-partition keys out of the path so the report
# surfaces the active embedding config, bertopic config, and variant
# alongside the fit summary.
topic_info_path <- tar_read(topics_runpod)
topics_dir <- dirname(topic_info_path)
parts <- strsplit(topics_dir, "/", fixed = TRUE)[[1]]
leaf_emb_cfg <- sub("^config=", "", grep("^config=", parts, value = TRUE))
leaf_btopic <- sub("^bertopic=", "", grep("^bertopic=", parts, value = TRUE))
leaf_variant <- sub("^variant=", "", grep("^variant=", parts, value = TRUE))

ti <- arrow::read_parquet(topic_info_path)
tp <- arrow::open_dataset(file.path(topics_dir, "topics.parquet"))

n_works <- nrow(tp)
n_topics <- nrow(ti)
n_noise <- ti |>
  dplyr::filter(topic_id == -1L) |>
  dplyr::summarise(n = sum(n_corpus + n_keypapers)) |>
  dplyr::pull(n)
if (length(n_noise) == 0) {
  n_noise <- 0L
}
n_topics_real <- n_topics - as.integer(any(ti$topic_id == -1L))

tibble::tibble(
  metric = c(
    "Number of topics (incl. noise -1)",
    "Number of topics (excl. noise)",
    "Works assigned to a topic",
    "Works in noise cluster (-1)",
    "Noise share (%)"
  ),
  value = c(
    format(n_topics, big.mark = ","),
    format(n_topics_real, big.mark = ","),
    format(n_works - n_noise, big.mark = ","),
    format(n_noise, big.mark = ","),
    sprintf("%.2f", 100 * n_noise / n_works)
  )
) |>
  knitr::kable(
    align = c("l", "r"),
    caption = sprintf(
      "High-level summary of the BERTopic fit (embedding config: %s · bertopic config: %s · variant: %s).",
      leaf_emb_cfg,
      leaf_btopic,
      leaf_variant
    )
  )
```

## Topic-size distribution

How is the corpus distributed across topics? A healthy fit shows
many topics with roughly comparable sizes (heavy-tailed but not
single-blob). A *collapsed* fit shows one topic absorbing the bulk
of the corpus — easy to spot visually as one bar dwarfing the rest.

```{r}
#| label: fig-topic-sizes
# Placeholder: fig_topic_sizes is not yet wired as a tar_target — build
# inline from topic_info. Move to a target once the visualisation layer
# is finalised (see TODO_Visualisations.md §2).
ti |>
  dplyr::filter(topic_id >= 0L) |>
  dplyr::arrange(dplyr::desc(n_corpus)) |>
  dplyr::mutate(rank = dplyr::row_number()) |>
  ggplot2::ggplot(ggplot2::aes(x = rank, y = n_corpus)) +
  ggplot2::geom_col(fill = "#5B8FF9") +
  ggplot2::scale_y_log10(labels = scales::label_comma()) +
  ggplot2::labs(
    x = "Topic rank (by size)",
    y = "Number of works (log scale)",
    title = "Topic-size distribution"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

```{r}
#| label: tbl-topic-size-quantiles
ti |>
  dplyr::filter(topic_id >= 0L) |>
  dplyr::pull(n_corpus) |>
  quantile(probs = c(0, 0.05, 0.25, 0.5, 0.75, 0.95, 1)) |>
  tibble::enframe(name = "quantile", value = "topic size") |>
  dplyr::mutate(`topic size` = format(round(`topic size`), big.mark = ",")) |>
  knitr::kable(
    align = c("l", "r"),
    caption = "Quantiles of the topic-size distribution (noise excluded)."
  )
```

## Topic vocabulary

For each topic: rank by size, top-N c-TF-IDF terms, count of works,
count of keypapers in the topic, and the `is_relevant` flag
(`n_keypapers ≥ keypaper_threshold`).

```{r}
#| label: tbl-topics
tar_read(tbl_topics)
```

## UMAP scatter

2-D UMAP of a sample of the corpus, with keypapers overlaid in full.
Coordinates come from `viz_umap_coords` (sampled corpus, seed +
`umap_sample_size` from `input/config.yaml: viz`).

```{r}
#| label: fig-umap
#| fig-width: 10
#| fig-height: 8
tar_read(viz_umap_fig)
```

## UMAP with topic overlay

Same 2-D UMAP, points coloured by their `topic_id` from the 5-D
HDBSCAN fit. Keypapers shown as larger markers in their own topic
colour.

```{r}
#| label: fig-umap-topics
#| fig-width: 10
#| fig-height: 8
tar_read(viz_topics_fig)
```

<!--
## Topic boundaries (density + concave hulls)

Disabled until the BERTopic fit is retuned — the current 6-topic
collapse would render as one giant polygon. See
TODO_Visualisations.md §"Wiring backlog".

```{r}
#| label: fig-umap-clusters
#| fig-width: 10
#| fig-height: 8
#| eval: false
tar_read(viz_umap_clusters_fig)
```
-->

# Keypaper coverage

Per keypaper: which topic does it land in, and does that topic clear
the `keypaper_threshold` (`is_relevant = TRUE`)? Topics that capture
≥ `keypaper_threshold` keypapers are treated as on-corpus topics for
downstream filtering.

```{r}
#| label: tbl-keypapers-by-topic
kp_topics <- arrow::open_dataset(file.path(topics_dir, "topics.parquet")) |>
  dplyr::filter(source == "keypaper") |>
  dplyr::select(id, topic_id, topic_source) |>
  dplyr::collect()

# Bring in keypaper titles directly from key_works.
kc <- arrow::open_dataset(tar_read(key_works)) |>
  dplyr::select(id, title) |>
  dplyr::collect() |>
  dplyr::mutate(citation = title)

kp_topics |>
  dplyr::left_join(kc, by = "id") |>
  dplyr::left_join(
    ti |> dplyr::select(topic_id, topic_name = label, is_relevant),
    by = "topic_id"
  ) |>
  dplyr::transmute(
    topic_id,
    topic = topic_name,
    relevant = ifelse(
      isTRUE(is_relevant) | is_relevant == TRUE,
      "<span style='color:green'>yes</span>",
      "<span style='color:red'>no</span>"
    ),
    source = topic_source,
    citation,
    title
  ) |>
  dplyr::arrange(topic_id) |>
  knitr::kable(
    format = "html",
    escape = FALSE,
    caption = "Topic assignment for each keypaper."
  )
```

```{r}
#| label: tbl-relevant-topic-summary
kp_topics |>
  dplyr::left_join(
    ti |> dplyr::select(topic_id, is_relevant, n_corpus),
    by = "topic_id"
  ) |>
  dplyr::summarise(
    keypapers = dplyr::n(),
    `works in same topics` = sum(n_corpus, na.rm = TRUE),
    .by = is_relevant
  ) |>
  dplyr::mutate(
    is_relevant = ifelse(is_relevant, "relevant topic", "off-corpus topic"),
    `works in same topics` = format(`works in same topics`, big.mark = ",")
  ) |>
  knitr::kable(
    align = c("l", "r", "r"),
    caption = "Keypaper distribution across relevant vs off-corpus topics."
  )
```

# Diagnostics

## Cluster collapse check

A single topic absorbing ≫ half the corpus is a strong indicator of
either UMAP collapsing into one blob or HDBSCAN's
`min_cluster_size` being too large for the embedding density. The
table below flags collapse if any topic holds ≥ 50% of the works.

```{r}
#| label: tbl-collapse-check
top_share <- ti |>
  dplyr::filter(topic_id >= 0L) |>
  dplyr::slice_max(n_corpus, n = 5) |>
  dplyr::mutate(
    share = n_corpus / sum(ti$n_corpus, na.rm = TRUE),
    share = sprintf("%.2f%%", 100 * share),
    n_corpus = format(n_corpus, big.mark = ",")
  ) |>
  dplyr::select(topic_id, label, n_corpus, share)
top_share |>
  knitr::kable(
    align = c("r", "l", "r", "r"),
    caption = "Top 5 topics by size — share of the corpus they absorb."
  )
```

## Primary vs fallback assignments

How many works are clustered via the primary (title+abstract) variant
vs the title-only fallback?

```{r}
#| label: tbl-topic-source
arrow::open_dataset(file.path(topics_dir, "topics.parquet")) |>
  dplyr::count(topic_source) |>
  dplyr::collect() |>
  dplyr::arrange(dplyr::desc(n)) |>
  dplyr::mutate(
    `%` = sprintf("%.2f", 100 * n / sum(n)),
    n = format(n, big.mark = ",")
  ) |>
  knitr::kable(
    align = c("l", "r", "r"),
    caption = "Assignment source per work (primary fit vs fallback projection)."
  )
```

# Provenance

```{r}
#| label: provenance
tibble::tibble(
  item = c(
    "Topics leaf",
    "BERTopic config name",
    "Embedding config",
    "Primary variant",
    "Fallback variant"
  ),
  value = c(
    topics_dir,
    leaf_btopic,
    leaf_emb_cfg,
    leaf_variant,
    cfg$fallback_variant %||% "title"
  )
) |>
  knitr::kable(
    caption = "Pointers to the on-disk artefacts behind this report."
  )
```
