---
title: "TCAC 2.0 — Vectorisation Report"
subtitle: "SPECTER2 embeddings, keypaper-similarity scoring"
author:
  - name:
      family: Krug
      given: Rainer M.
    id: rmk
    orcid: 0000-0002-7490-0066
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 5
    grid:
        sidebar-width: 0px
        body-width: 4000px
        margin-width: 200px
        gutter-width: 1.5rem      
execute:
  message: false
  warning: false
  cache: false
---

```{r}
#| label: setup
library(targets)
library(htmltools)
library(knitr)
```

# Run metadata

```{r}
#| label: metadata
emb_meta <- tar_read(viz_metadata)
model_id <- tar_read(viz_embeddings) |>
  dplyr::distinct(model_id) |>
  dplyr::pull()
emb_meta |>
  knitr::kable(
    caption = sprintf(
      "Embedded works per source × variant (model_id = %s).",
      model_id
    )
  )
```

# Score distributions per variant

For each corpus work, the **max** cosine similarity against any keypaper is the
single most useful summary — it answers "how close is this work to its nearest
reference paper?" A well-calibrated embedding produces a bimodal distribution:
relevant works pile up on the right, irrelevant on the left.

```{r}
#| label: tbl-score-summary
tar_read(viz_score_summary_tbl)
```

```{r}
#| label: fig-score-dist
#| fig-cap: "Distribution of max cosine similarity to keypapers, per variant. Higher = closer to nearest reference. A clear bimodal shape indicates the embedding separates topic-relevant from off-topic works."
tar_read(viz_score_dist)
```

Median and quantile breakdown:

```{r}
#| label: tbl-score-quantiles
tar_read(viz_score_quantiles_tbl) |>
  knitr::kable(digits = 3)
```

# Top / bottom matches per variant

Sanity check: do the highest-scoring works read as on-topic, and the lowest as
off-topic? If not, either the embedding model isn't capturing the topic, the
keypaper set is too narrow, or the search-term filter let in noise.

```{r}
#| label: tbl-top-bottom
#| results: asis
tb <- tar_read(viz_top_bottom)
for (v in names(tb)) {
  cat("##", v, "\n\n")
  print(knitr::kable(tb[[v]]))
  cat("\n\n")
}
```

# Variant agreement

If `title` and `abstract` agree on a work's relevance, the title is a reliable
proxy for the abstract. Off-diagonal works are interesting candidates for
manual inspection: either the title misrepresents the paper or the abstract is
poor quality.

```{r}
#| label: fig-variant-agree
#| fig-cap: "Per-work max similarity to keypapers, plotted pair-wise across the three variants. Each corpus work contributes one point per pair (coloured). The dashed diagonal is perfect agreement; points off it are works the two variants disagree about. Click or box/lasso-select to highlight a work's points across all pairs."
tar_read(viz_agree_fig)
```

# Threshold

```{r}
#| label: fig-threshold-sweep
#| fig-cap: "Number of corpus works with max cosine ≥ threshold, per variant."
tar_read(viz_threshold)
```

# Topic-space map (UMAP)

Drag the slider underneath the map to highlight all corpus works whose overall
max cosine to any keypaper meets or exceeds the threshold.

UMAP projection of the `title_abstract` embeddings for corpus + keypapers.
Corpus points are coloured by their max-similarity to the keypaper set, so
clusters of high-relevance works should pile up near the keypapers (marked
in red).

Interactive map (hover for the citation, title, and max cosine; single-click
on a keypaper or corpus point to highlight its cluster and draw connecting
lines; corpus colour is the work's **max cosine to any keypaper**):

```{r}
#| label: contour-toggle
#| results: asis
toggle <- htmltools::tagList(
  htmltools::tags$div(
    style = "margin: 0.5em 0;",
    htmltools::tags$input(
      type = "checkbox",
      id = "contour_toggle",
      checked = NA
    ),
    htmltools::tags$label(
      `for` = "contour_toggle",
      " Show similarity contour underlay"
    )
  ),
  htmltools::tags$script(htmltools::HTML(
    "(function () {
       var toggle = document.getElementById('contour_toggle');
       toggle.addEventListener('change', function () {
         if (window._tcacUmapEl) {
           Plotly.restyle(window._tcacUmapEl, { visible: toggle.checked }, [0]);
         }
       });
     })();"
  ))
)
```

```{r}
#| label: threshold-slider
#| results: asis
slider <- htmltools::tagList(
  htmltools::tags$div(
    style = "display:flex; gap:1em; align-items:center; margin: 0.5em 0 1.5em 0;",
    htmltools::tags$label(`for` = "threshold_slider", "Threshold:"),
    htmltools::tags$input(
      type = "range",
      id = "threshold_slider",
      min = "0",
      max = "1",
      step = "0.01",
      value = "0",
      style = "flex: 1;"
    ),
    htmltools::tags$span(
      id = "threshold_value",
      "0.00",
      style = "min-width: 3em; text-align: right; font-family: monospace;"
    )
  ),
  htmltools::tags$script(htmltools::HTML(
    "(function () {
       var slider  = document.getElementById('threshold_slider');
       var display = document.getElementById('threshold_value');
       slider.addEventListener('input', function (e) {
         var t = parseFloat(e.target.value);
         display.textContent = t.toFixed(2);
         if (window.tcacSetThreshold) window.tcacSetThreshold(t);
       });
     })();"
  ))
)
```

::: {style="border:1px solid #ccc; border-radius:6px; padding:1em; background:#fafafa; margin:1em 0;"}

```{r}
#| label: show_map
#| echo: false
#| fig-height: 9
toggle
tar_read(viz_umap)
slider
```

:::

# Topics

BERTopic clusters of the SPECTER2 `title_abstract` embeddings (no-abstract
works are transferred via their `title` embedding to the nearest cluster). See
[TD_BERTopic.md](TD_BERTopic.md). Topics are sorted by **number of keypapers**
in the cluster — clusters at the top are the most relevant to TCAC's reference
set.

```{r}
#| label: tbl-topics
tbl <- tar_read(viz_topics_tbl) |>
  DT::formatStyle(
    columns = 6, # column index (1-based here)
    fontSize = '60%' # or '80%', '0.8em', etc.
  )

# Add columnDefs to the existing datatable object
tbl$x$options$columnDefs <- list(
  list(width = '600px', targets = -1),
  list(width = '400px', targets = -2)
)
tbl$x$options$pageLength <- 50

# Also ensure autoWidth is enabled so your widths are respected
tbl$x$options$autoWidth <- TRUE

tbl

rm(tbl)
```

```{r}
#| label: fig-umap-topics
#| fig-height: 9
#| fig-cap: "UMAP coloured by BERTopic cluster id. Clustering ran on the 5-D UMAP (not this 2-D view), so spatial proximity here is only loosely correlated with cluster membership. Keypapers (red triangles) overlay the corpus."
tar_read(viz_topics)
```
