---
title: 'Transformative Change Assessment Corpus 2.0'
subtitle: 'Extending TCAC 1.0 for added functionality'
author:
  - name: 
        family: Krug
        given: Rainer M.    
    id: rmk
    orcid: 0000-0002-7490-0066
    email: Rainer.Krug@sib.swiss, Rainer@Krugs.de
    roles: [author, editor] 
  - name: 
        family: Villasante
        given: Sebastian    
    id: sv
    orcid: 0000-0001-6296-4479
    roles: [author] 

abstract: > 
 
license: "CC BY"

doi: &doi XXXXXXX
citation: 
  type: report
  container-title: 'TCA Corpus 2.0'
  doi: *doi
version: 2.0.0 

format:
    html:
        toc: true
        toc-depth: 5
        toc_expand: true
        embed-resources: true
        code-fold: true
        code-summary: 'Show the code'
        link-external-newwindow: true
        grid:
            sidebar-width: 0px
            body-width: 4000px
            margin-width: 200px
            gutter-width: 1.5rem      
execute:
  message: NA

params:
   

---


```{r}
#| label: setup

library(targets)
library(dplyr)
library(arrow)
```

# Introduction

## Data Base
The data base is the OpenAlex snapshot from the `RELEASE 2026-01-15`. All metadata is extracted from this snapshot, based on the openalex ids as identified for the TCAC 1.0 and newly identified for the TCAC 2.0. This guarantees consistency in the metadata and avoids any bias based on updated metadata.


# Methods

## Functions Underlying the Compilation and Analysis of the Corpora

## Get Key Paper

The DOIs of the key paper used for assessing the inclusiveness of TCAC 1.0 are used here and the works are downloaded

```{r}
#| label: show-prepare-keypapers
#| eval: false
#| file: "R/prepare_keypapers.R"
#| code-fold: true
#| code-summary: "Show prepare_keypaper.R"
```

## Get Corpora from Snapshot based on IDs

For consistency, both corpora are extracted, using the OpenAlex ids, from the OpenAlex snapshot `RELEASE 2026-01-15`. 


```{r}
#| label: show-get_corpus_from_snapshot
#| eval: false
#| file: "R/get_corpus_from_snapshot.R"
#| code-fold: true
#| code-summary: "Show get_corpus_from_snapshot.R"
```


## Count of works currently

The number of works on OpenAlex is determined using the following search terms

- Transformative change search term [tfc.txt](input/search terms/tfc_TCAC_1.0.txt){target="_blank"}
- Nature search term [nature.txt](input/search terms/nature_TCAC_1.0.txt){target="_blank"}

combined with `AND`as well as

- Types Filter [openalex_types.csv](input/openalex_types.csv){target="_blank"}

```{r}
#| label: show-get_count
#| eval: false
#| file: "R/get_count.R"
#| code-fold: true
#| code-summary: "Show get_count.R"
```

## Download of TCAC 2.0 ids

To make the download as efficient as possible as well as to keep consistency in the metadata with the above described TCAC 1.0, the ids of TCAC 2.0 are downloaded from OpenAlex and the full records are retrieved from the same snapshot as TCAC 1.0 has been re-constructed.


```{r}
#| label: show-get_tcac20_ids
#| eval: false
#| file: "R/get_tcac20_ids.R"
#| code-fold: true
#| code-summary: "Show get_tcac20_ids.R"
```

The extraction of the actual TCAC 2.0 is done again using the function as described above.

# Results

