---
title: 'Transformative Change Assessment Corpus 2.0'
subtitle: 'Extending TCAC 1.0 for added functionality'
author:
  - name: 
        family: Krug
        given: Rainer M.    
    id: rmk
    orcid: 0000-0002-7490-0066
    email: Rainer.Krug@sib.swiss, Rainer@Krugs.de
    roles: [author, editor] 
  - name: 
        family: Villasante
        given: Sebastian    
    id: sv
    orcid: 0000-0001-6296-4479
    roles: [author] 

abstract: > 
 
license: "CC BY"

doi: &doi XXXXXXX
citation: 
  type: report
  container-title: 'TCA Corpus 2.0'
  doi: *doi
version: 2.0.0 

format:
    html:
        toc: true
        toc-depth: 5
        toc_expand: true
        embed-resources: true
        code-fold: true
        code-summary: 'Show the code'
        link-external-newwindow: true
        grid:
            sidebar-width: 0px
            body-width: 4000px
            margin-width: 200px
            gutter-width: 1.5rem      
execute:
  message: NA

params:
   

---


```{r}
#| label: setup
library(targets)
library(dplyr)
library(arrow)
```

# Introduction

## Data Base
The data base is the OpenAlex snapshot from the `RELEASE 2026-01-15`. All metadata is extracted from this snapshot, based on the openalex ids as identified for the TCAC 1.0 and newly identified for the TCAC 2.0. This guarantees consistency in the metadata and avoids any bias based on updated metadata.


# Methods


![TCAC 2.0 pipeline overview as a directed acyclic graph, grouped into four stages: OpenAlex corpus assembly (search-term query → IDs → snapshot extraction), per-variant SPECTER2 embedding (via local or RunPod TEI), BERTopic clustering (Path A laptop CPU, Path B RunPod GPU reading from Cloudflare R2), and the visualisation layer that feeds the Quarto report. Shapes encode artefact type: documents for input files, cylinders for databases (local or cloud), rectangles for processing targets, and clouds for external services.](output/figures/mmd/pipeline.svg){#fig-pipeline-dag fig-alt="Directed acyclic graph of the TCAC 2.0 pipeline showing four subgraphs (corpus assembly, embeddings, topics dual-path, and visualisation/report) connected by directed edges between input files, databases, processing targets, and external services."}

![End-to-end TCAC 2.0 pipeline sequence: the laptop orchestrator drives OpenAlex corpus assembly, per-variant SPECTER2 embedding (local TEI or RunPod TEI), optional Cloudflare R2 mirror, BERTopic clustering (Path A laptop CPU vs Path B RunPod GPU reading from R2), and final Quarto report generation.](output/figures/mmd/sequence.svg){#fig-pipeline-sequence fig-alt="Sequence diagram of the TCAC 2.0 pipeline showing the orchestrator coordinating OpenAlex API, local snapshot, local or RunPod TEI embedding, Cloudflare R2 storage, local or RunPod BERTopic clustering, and Quarto report rendering across five stages."}

The pipeline orchestrator runs on the laptop and dispatches every step via R's `targets` package. Stage 1 assembles the corpus by querying the OpenAlex REST API for matching IDs and resolving keypapers, then extracting full records from a local OpenAlex snapshot. Stage 2 generates per-variant SPECTER2 embeddings via a TEI inference server — either local CPU/GPU or hosted on a RunPod GPU pod. Stage 3 mirrors the resulting parquet leaves to Cloudflare R2 (S3-compatible object storage), which decouples the laptop from the cluster-compute step. Stage 4 runs BERTopic on the embeddings, in one of two variants: a laptop-CPU sample-fit-and-transfer (Path A) for fast iteration, or a RunPod GPU full-corpus fit (Path B) where the pod reads embeddings directly from R2 via duckdb's `httpfs` extension. Stage 5 scores corpus works against keypapers, prepares visualisation data (`viz_*` targets), renders figures (`fig_*` targets), and assembles the Quarto report.

The DAG view (@fig-pipeline-dag) emphasises data flow and artefact types; the sequence view (@fig-pipeline-sequence) emphasises the temporal order of interactions across actors during a single `tar_make()` invocation. The two are complementary — most readers will find the DAG sufficient for understanding *what* the pipeline does, and the sequence diagram useful for understanding *how* the orchestrator coordinates the external services.





## Functions Underlying the Compilation and Analysis of the Corpora

## Get Key Paper

The DOIs of the key paper used for assessing the inclusiveness of TCAC 1.0 are used here and the works are downloaded

```{r}
#| label: show-prepare-keypapers
#| eval: false
#| file: "R/prepare_keypapers.R"
#| code-fold: true
#| code-summary: "Show prepare_keypaper.R"
```

## Get Corpora from Snapshot based on IDs

For consistency, both corpora are extracted, using the OpenAlex ids, from the OpenAlex snapshot `RELEASE 2026-01-15`. 


```{r}
#| label: show-get_corpus_from_snapshot
#| eval: false
#| file: "R/get_corpus_from_snapshot.R"
#| code-fold: true
#| code-summary: "Show get_corpus_from_snapshot.R"
```


## Count of works currently

The number of works on OpenAlex is determined using the following search terms

- Transformative change search term [tfc.txt](input/search terms/tfc_TCAC_2.0.txt){target="_blank"}
- Nature search term [nature.txt](input/search terms/nature_TCAC_2.0.txt){target="_blank"}

combined with `AND`as well as

- Types Filter [openalex_types.csv](input/openalex_types.csv){target="_blank"}

```{r}
#| label: show-get_count
#| eval: false
#| file: "R/get_count.R"
#| code-fold: true
#| code-summary: "Show get_count.R"
```

## Download of TCAC 2.0 ids

To make the download as efficient as possible as well as to keep consistency in the metadata with the above described TCAC 1.0, the ids of TCAC 2.0 are downloaded from OpenAlex and the full records are retrieved from the same snapshot as TCAC 1.0 has been re-constructed.


```{r}
#| label: show-get_tcac20_ids
#| eval: false
#| file: "R/get_tcac20_ids.R"
#| code-fold: true
#| code-summary: "Show get_tcac20_ids.R"
```

The extraction of the actual TCAC 2.0 is done again using the function as described above.

# Results

